First Order Alternation

### 2018-10-05 quantifiers/draft.tex
ok

\begin{definition}\label{def:coverage}
  Given an unfolding $U$ of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ a node $\alpha \in \dom(U)$ is
  \emph{covered} by another node $\beta \in \dom(U)$, denoted $\alpha
  \cover \beta$, if and only if there exists a node $\alpha' \prefix
  \alpha$ such that, for all words $w \in \Sigma[X]^*$ and all cubes
  $c \in \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$
  then there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that
  $\A$ accepts $w$ started in $d$. Moreover, $U$ is \emph{closed} if
  and only if every leaf from $\dom(U)$ is covered by an uncovered
  node.
\end{definition}

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. For a detailed presentation of the emptiness procedure, we
refer the reader to \cite{IosifXu18} (or Appendix
\ref{app:pseudocode}).

The semi-algorithm stops when (and if) a safe complete unfolding has
been found. The following theorem proves the soundness of this
termination argument:

\begin{theorem}\label{thm:soundness}
  If an automaton $\A$ has a nonempty safe closed unfolding then
  $\lang{A} = \emptyset$.
\end{theorem}
\proof{ 
  Let $U$ be a safe and complete unfolding of $\A$, such that
  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
  exists a word $w \in \lang{A}$ and let $\alpha \isdef
  \event{w}$.

  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, there
  exists an interpretation $\I$ such that $\I,\data{w} \models
  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
  case, one can show, by induction on the length $n\geq0$ of $w$, that
  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
  contradiction.

  We proceed thus under the assumption that $\alpha \not\in
  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
  $\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered
  and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
  that, for some uncovered node $\beta_1 \in \dom(U)$, for every word
  $w'$ and every cube $c \in \cube{\sem{U(\alpha_1)}}$ such that $\A$
  accepts $w'$ starting with $c$, there exists a cube $d \in
  \cube{\sem{U(\beta_1)}}$ such that $\A$ accepts $w'$ starting with
  $d$. Let $\gamma_1$ be the unique sequence such that
  $\alpha_1\gamma_1 = \alpha$. By Definition \ref{def:coverage}, since
  $\alpha_1 \cover \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in
  \lang{\A}$, there exists a word $w_1$ and a cube $c_1 \in
  \cube{\sem{U(\beta_1)}}$ such that $\event{w_1} = \gamma_1$ and $\A$
  accepts $w_1$ starting with $c_1$. If $\beta_1\gamma_1 \in \dom(U)$,
  we obtain a contradiction by a similar argument as above. Hence
  $\beta_1\gamma_1 \not\in \dom(U)$ and there exists a leaf of
  $\dom(U)$ which is also a prefix of $\beta_1\gamma_1$. Since $U$ is
  closed, this leaf is covered by an uncovered node $\beta_2 \in
  \dom(U)$ and let $\alpha_2 \in \dom(U)$ be the minimal (in the
  prefix partial order) node such that $\beta_1 \prefix \alpha_2
  \prefix \beta_1\gamma_1$ and $\alpha_2 \cover \beta_2$. Let
  $\gamma_2$ be the unique sequence such that $\alpha_2\gamma_2 =
  \beta_1\gamma_1$. Since $\beta_1$ is uncovered, we have $\beta_1
  \neq \alpha_2$ and thus $\len{\gamma_1} > \len{\gamma_2}$. By
  repeating the above reasoning for $\alpha_2$, $\beta_2$ and
  $\gamma_2$, we obtain an infinite sequence $\len{\gamma_1} >
  \len{\gamma_2} > \ldots$, which is again a contradiction. \qed}

@@

  \noindent(\ref{it2:subst-quant}) With the notation introduced at
  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
  \tau_{i_\ell}$, whose existence is proved by point
  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi$ is
  monotonic, we obtain: \begin{compactitem}
  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
    \stamp{X}{0} \cup \set{x_j \mid j < \max(\xi^{-1}(0)), Q_j =
      \exists}$,
    %
  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
    %
  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
  \end{compactitem}
  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
        \max(\xi^{-1}(0)), Q_j = \exists}$,
      %
    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
      %
    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
  \end{compactitem}
  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
      Theorem \ref{thm:skolem},
      %
    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
      \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1}\Big) \models
      I_{k-1} \wedge \phi_k \models I_k$, and 
      %
    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
      \models I_n \wedge \psi \models \bot$,     
  \end{compactitem}
  as required by Definition
  \ref{def:generalized-lyndon-interpolant}. \qed}

@@

Suppose that a given input event sequence $\alpha \in \dom(U)$ is
labeled with a sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1),
\ldots U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement
from Definition \ref{def:unfolding} (\ref{it3:unfolding}) and,
moreover, it has a non-local GLI $(I_0, \ldots, I_n)$, such that
$\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
renamed variables occurring in the rules of $\Delta$ under the scope
of existential transition quantifiers (formally these are the
existentially quantified variables from the prenex normal form of
$\quantform{\alpha_k}$). The interpolant is used in refining the
labeling of the sequence as follows, for all $k \in [n]$: \begin{compactitem}
\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
  from $I_k$ by\begin{inparaenum}[(i)]
    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
      %
    \item existentially quantifying each variable $y \in
      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$.
  \end{inparaenum}
%
\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
\end{compactitem}
To see that the new set of formulae $U'(\alpha_0), \ldots,
U'(\alpha_n)$ complies with Definition \ref{def:unfolding}, it is
enough to notice the following facts: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}.} $\stamp{\iota}{0} \models
  I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) = U(\alpha_0)
  \wedge J_0 \equiv \iota \wedge J_0$.
%
\item 
\end{compactitem}

%% \subsection{Coverage}

%% Assuming that the data domain is countable, the set of finite words
%% $\Sigma[X]^*$ is countable and the emptiness problem is recursively
%% enumerable. Moreover, by Lemma \ref{lemma:acceptance}, a systematic
%% enumeration of sequences of input events $\alpha \in \Sigma^*$ will
%% eventually discover an accepted word, provided that all path formulae
%% $\pathform{\alpha}$ belong to a decidable first-order theory.

%% However, if the automaton is empty, the enumeration of input event
%% sequences will fail to terminate, in general. The classical way to
%% fight this divergence problem is to consider a \emph{coverage} partial
%% order, which is a subset of the following relation:

%% \begin{definition}\label{def:coverage}
%%   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and
%%   $S_1, S_2 \subseteq \bigcup_{q \in Q} \set{q(d_1,\ldots,d_{\#(q)})
%%     \mid d_1,\ldots,d_{\#(q)} \in \Data}$, we have $S_1 \sqsubseteq
%%   S_2$ if and only if 
%%   \begin{equation}\label{eq:po}
%%     \text{
%%       \begin{minipage}{0.7\textwidth}
%%         $\A$ accepts a word $w$ when started in a set of cubes $C_1
%%         \in S_1$ only if $\A$ accepts $w$ when started in a set of
%%         cubes $C_2 \in S_2$.
%%     \end{minipage}} 
%%   \end{equation}
%% \end{definition}
%% Since, in general the $\sqsubseteq$ relation is not definable using
%% available decidable (SMT) theories, we consider the following stronger
%% relation: $S_1 \supsetforex S_2$ if and only if, for each set of cubes
%% $C_1 \in S_1$, there exists a set of cubes $C_2 \in S_2$ such that
%% $C_1 \supseteq C_2$.

%% \begin{lemma}\label{lemma:coverage}
%%   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
%%   sets of cubes $S_1, S_2 \subseteq \bigcup_{q \in Q}
%%   \set{q(d_1,\ldots,d_{\#(q)}) \mid d_1,\ldots,d_{\#(q)} \in \Data}$
%%   such that $S_1 \supsetforex S_2$, we have $S_1 \sqsubseteq S_2$.
%% \end{lemma}
%% \proof{  
%% \qed}

>>>>

\begin{definition}\label{def:coverage}
  Given an unfolding $U$ of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ a node $\alpha \in \dom(U)$ is
  \emph{covered} by another node $\beta \in \dom(U)$, denoted $\alpha
  \cover \beta$, if and only if there exists a node $\alpha' \prefix
  \alpha$ such that $U(\alpha') \models U(\beta)$. Moreover, $U$ is
  \emph{closed} if and only if every leaf from $\dom(U)$ is covered by
  an uncovered node.
\end{definition}

%% , for all words $w \in \Sigma[X]^*$ and all cubes $c \in
%% \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$ then
%% there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that $\A$
%% accepts $w$ started in $d$.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer the reader to \cite{IosifXu18} (or Appendix
\ref{app:pseudocode}).

\begin{theorem}\label{thm:soundness}
  If an automaton $\A$ has a nonempty safe closed unfolding then
  $\lang{A} = \emptyset$.
\end{theorem}
\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
  exists a word $w \in \lang{A}$ and let $\alpha \isdef
  \event{w}$. Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance},
  there exists an interpretation $\I$ such that $\I,\data{w} \models
  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
  case, one can show, by induction on the length $n\geq0$ of $w$, that
  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
  contradiction.

  We proceed thus under the assumption that $\alpha \not\in
  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
  $\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered
  and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
  that $U(\alpha_1) \models U(\beta_1)$, for some uncovered node
  $\beta_1 \in \dom(U)$.
  %% for every word $w'$ and every cube $c \in
  %% \cube{\sem{U(\alpha_1)}}$ such that $\A$ accepts $w'$ starting
  %% with $c$, there exists a cube $d \in \cube{\sem{U(\beta_1)}}$
  %% such that $\A$ accepts $w'$ starting with $d$.
  Let $\gamma_1$ be the unique sequence such that $\alpha_1\gamma_1 =
  \alpha$. By Definition \ref{def:coverage}, since $\alpha_1 \cover
  \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in \lang{\A}$, there
  exists a word $w_1$ and a cube $c_1 \in \cube{\sem{U(\alpha_1)}}
  \subseteq \cube{\sem{U(\beta_1)}}$, such that $\event{w_1} =
  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. If
  $\beta_1\gamma_1 \in \dom(U)$, we obtain a contradiction by a
  similar argument as above. Hence $\beta_1\gamma_1 \not\in \dom(U)$
  and there exists a leaf of $\dom(U)$ which is also a prefix of
  $\beta_1\gamma_1$. Since $U$ is closed, this leaf is covered by an
  uncovered node $\beta_2 \in \dom(U)$ and let $\alpha_2 \in \dom(U)$
  be the minimal (in the prefix partial order) node such that $\beta_1
  \prefix \alpha_2 \prefix \beta_1\gamma_1$ and $\alpha_2 \cover
  \beta_2$. Let $\gamma_2$ be the unique sequence such that
  $\alpha_2\gamma_2 = \beta_1\gamma_1$. Since $\beta_1$ is uncovered,
  we have $\beta_1 \neq \alpha_2$ and thus $\len{\gamma_1} >
  \len{\gamma_2}$. By repeating the above reasoning for $\alpha_2$,
  $\beta_2$ and $\gamma_2$, we obtain an infinite sequence
  $\len{\gamma_1} > \len{\gamma_2} > \ldots$, which is again a
  contradiction. \qed}

@@

  \noindent(\ref{it2:subst-quant}) With the notation introduced at
  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
  \tau_{i_\ell}$, whose existence is proved by point
  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi$ is
  monotonic, we obtain: \begin{compactitem}
  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
    \stamp{X}{0} \cup \set{x_j \mid j < \max(\xi^{-1}(0)), Q_j =
      \exists}$,
    %
  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
    %
  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
  \end{compactitem}
  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
        \max(\xi^{-1}(0)), Q_j = \exists}$,
      %
    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
      %
    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
  \end{compactitem}
  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
      Theorem \ref{thm:skolem},
      %
    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
      I_{k-1} \wedge \phi_k \models I_k$, and 
      %
    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
      \models I_n \wedge \psi \models \bot$,     
  \end{compactitem}
  as required by Definition
  \ref{def:generalized-lyndon-interpolant}. \qed}

@@

Suppose that a given input event sequence $\alpha \in \dom(U)$ is
labeled with a sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1),
\ldots U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement
from Definition \ref{def:unfolding} (\ref{it3:unfolding}) and,
moreover, it has a non-local GLI $(I_0, \ldots, I_n)$, such that
$\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
renamed variables occurring in the rules of $\Delta$ under the scope
of existential transition quantifiers (formally these are the
existentially quantified variables from the prenex normal form of
$\quantform{\alpha_k}$). The interpolant is used in refining the
labeling of the sequence as follows, for all $k \in [n]$: \begin{compactitem}
\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
  from $I_k$ by\begin{inparaenum}[(i)]
    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
      %
    \item existentially quantifying each variable $y \in
      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$.
  \end{inparaenum}
%
\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
\end{compactitem}

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of\begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data
values seen in the input at some earlier moment in time, or 
%
\item variables $\set{x_{i_1},\ldots,x_{i_k}}$ that track past choices
  made within the transition rules. 
\end{inparaenum}
From the present point of view, it is not important when exactly in
the past a certain input has been read or when a choice has been made,
but rather the (symbolic) value and its associated data
constraints. Intuitively, existential quantification of these
variables does the job of ignoring the particular moments in the past
when they have been encountered. Formally, to see that the new set of
formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
\ref{def:unfolding}, it is enough to notice the following
facts: \begin{compactitem}
\item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
  \ref{it2:generalized-lyndon-interpolant} of Definition
  \ref{def:generalized-lyndon-interpolant}.} $\stamp{\iota}{0} \models
  I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) = U(\alpha_0)
  \wedge J_0 \equiv \iota \wedge J_0$.
%
\item by Definition \ref{def:generalized-lyndon-interpolant}
  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
  [n-1]$:
  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
  \leftarrow \stamp{\psi}{k+1} \models I_{k+1}\] By slightly abusing
  notation, let $\stamp{I}{n}_\ell$ be the formula in which every
  predicate symbol $\stamp{q}{\ell}$ has been replaced with
  $\stamp{q}{n}$. Then the following entailment holds: 
  \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \leftarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
  Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \leftarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
  satisfies the labeling condition of Definition \ref{def:unfolding}
  (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
  we obtain, as required: 
  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
  \leftarrow \stamp{\psi}{1} \models
  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
\end{compactitem}

The last ingredient of the lazy annotation semi-algorithm based on
unfolding refinement with coverage consist in the implementation of
the coverage check, when the unfolding of an automaton is labeled with
conjunctions of existentially quantified formulae with predicate
symbols, obtained from interpolation. By Definition
\ref{def:coverage}, checking whether a given node $\alpha \in \dom(U)$
is covered amounts to finding a prefix $\alpha' \prefix \alpha$ and a
node $\beta \in \dom(U)$ such that $U(\alpha') \models U(\beta)$, or
equivalently, the formula $U(\alpha') \wedge \neg U(\beta)$ is
unsatisfiable. However, the latter formula, in prenex form, has
quantifier prefix in the language $\exists^*\forall^*$ and, as
previously mentioned, the satisfiability problem for such formulae
becomes undecidable when the data theory subsumes Presburger
arithmetic \cite{Halpern91}.

Nevertheless, if we require just a yes/no answer (i.e.\ not an
interpolant) recently developped quantifier instantiation heuristics,
such as \cite{ReynoldsKK17}, perform rather well in answering a large
number of queries in this class. Observe, moreover, that coverage does
not need to rely on a complete decision procedure. If the prover fails
in answering the above satisfiability query, then the semi-algorithm
assumes the node is not covered and continues exploring its
successors. Failure to compute complete coverage may lead to
divergence (non-termination) and failure to prove emptiness, but the
soundness of the procedure is preserved.
